Stimjim is open source hardware for precise electrical stimulation

### 2019-06-05 benchmarking/paper/draft2.Rnw
added figure captions, more references to analysis.

\documentclass[twocolumn]{article}
%\documentclass{article}
\usepackage[margin=2cm]{geometry}
%\usepackage[superscript,biblabel]{cite}
\usepackage[super]{natbib}
\begin{document}
%\SweaveOpts{concordance=TRUE}

@@

    ishe = which(mmand::erode(mmand::erode(isStageHigh>0, c(rep(1,61),rep(0,60))), rep(1,61)) > 0)
    isle = which(mmand::erode(mmand::erode(isStageLow>0, c(rep(1,61),rep(0,60))), rep(1,61)) > 0)
    sh.median = mean(x[ishe])*vpau
    sh.sd = sd(x[ishe])*vpau
    sl.median = mean(x[isle])*vpau
    sl.sd = sd(x[isle])*vpau

    pulseMat = sapply(pulseStartIndices, function(j) x[j+(-100:1000)])
    list(IPI.summary=IPI.summary, IPI.sd=IPI.sd, PW.summary=PW.summary, 
             PW.sd=PW.sd, sh.median=sh.median, sh.sd=sh.sd, 
             sl.median=sl.median, sl.sd=sl.sd, pulseMat=pulseMat)
  })
}

@@

\section*{Abstract}
Electrical stimulation is a simple and powerful tool to perturb and understand the function of neurons and neural circuits. Despite this, devices that can provide precise current or voltage stimulation are expensive and closed-source. Here, we introduce Stimjim, a capable and inexpensive (\$200) open-source instrument for electrical stimulation that combines both function generation and electrical isolation. Stimjim provides microsecond temporal resolution with microampere or millivolt level precision on two electrically isolated output channels. To show the utility of Stimjim, we use it to train mice to perform intracranial self-stimulation (ICSS) for brain stimulation reward (BSR) in a novel head-fixed paradigm. We show that reward-seeking behavior can be smoothly tuned by adjusting either stimulation frequency or amplitude.

@@

In contrast, there has been a recent push within the scientific community to produce open labware - open source hardware and software replacements for a variety of common laboratory tasks. Examples include Open Ephys\cite{siegle_open_2017}, Cyclops\cite{newman_optogenetic_2015}, OpenPCR\cite{noauthor_openpcr_nodate}, FlyPi\cite{chagas_100_2017}, Openspritzer\cite{forman_openspritzer:_2017}, and Audiomoth\cite{hill_audiomoth:_2018}. We now add Stimjim to this growing body of open hardware. Stimjim replaces \textit{in vivo} neural stimulators at a fraction of the cost, without compromising on quality. Furthermore, due to its entirely open design and software, Stimjim can be modified by end users to fit their specific needs.

@@

<<benchmarkFigure, echo=FALSE,fig.height=8,fig.width=5,fig.cap="\\label{fig:benchmarking} \\textbf{Electrical benchmarks show Stimjim provides microsecond temporal precision and millivolt- and microampere-scale amplitude precision.}", fig.align='center', out.width='1\\linewidth', fig.env="figure">>=

@@

plot(paramMat[1:30,3]/1e3, sapply(res[seq(1,60,by=2)], function(x) x$IPI.sd)*1e6, yaxt='n',
  log='x', main='', cex=0.75, xlab='Target IPI (ms)', ylab='IPI standard deviation (us)')
points(paramMat[1:30,3]/1e3, sapply(res[seq(2,60,by=2)], function(x) x$IPI.sd)*1e6,
  col='red', cex=0.75)
axis(2,las=2)
mtext("C",side=2, padj=padj,las=2)
# 
# plot(paramMat[1:30,3], sapply(res[seq(1,60,by=2)], function(x) diff(x$IPI.summary[c(1,6)]))*1e6, 
#   log='x', main='IPI range', cex=0.75, xlab='Target IPI', ylab='Range of measured IPIs')
# points(paramMat[1:30,3], sapply(res[seq(2,60,by=2)], function(x) diff(x$IPI.summary[c(1,6)]))*1e6,
#   col='red', cex=0.75)

@@

plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) x$PW.sd)*1e6/2, yaxt='n',
  log='x', main='', xlab='Target PW (us)', ylab='PW standard deviation (us)', cex=0.75)
points(paramMat[31:59,7], sapply(res[seq(62,118,by=2)], function(x) x$PW.sd)*1e6/2,
  col='red', cex=0.75)
axis(2,las=2)
mtext("F",side=2, padj=padj,las=2)
# plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) diff(x$PW.summary[c(1,6)]))*1e6/2, 
#   log='x', main='PW range', ylab='PW range', cex=0.75)
# points(paramMat[31:59,7], sapply(res[seq(62,118,by=2)], function(x) diff(x$PW.summary[c(1,6)]))*1e6/2,
#   col='red', cex=0.75)

ex3 = sapply(dataFiles[seq(81,61,length.out=6)], function(f){
  dat = matrix(readBin(f, what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)
  fp = min(which(dat[,1] > 10))
  dat[fp + (-200:700),1]*vpau
})
par(mar=c(4,1,2,1))
matplot(1e3/fs*1:nrow(ex3), dendRites::offset(ex3,15), col=1, type='l', lty=1, bty='n', yaxt='n', ylab='', xlab='time (ms)',
                main = "Varying amplitude");
mtext("G",side=2, padj=padj,las=2)
par(mar=c(4,3,2,1))
x = c(paramMat[61:89,5], -paramMat[61:89,5])
y = t(sapply(res[seq(121,178,by=2)], function(x) c(x$sh.median,x$sl.median))*1e3) * 0.458/0.4593
plot(x/1000,y-x, main='Voltage accuracy', xlab='Target amplitude (V)', ylab='Measured error (mV)', cex=0.75, yaxt='n', ylim=c(-30,30))
axis(2,las=2)
mtext("H",side=2, padj=padj,las=2)

x2 = c(paramMat[61:90,6], -paramMat[61:90,6])
y2 = t(sapply(res[seq(122,180,by=2)], function(x) c(x$sh.median,x$sl.median))/9750*1e6)
plot(x2/1000,y2-x2, main='Current accuracy', xlab='Target amplitude (mA)', ylab='Measured error (uA)', cex=0.75, col='red', yaxt='n',ylim=c(-5,5))
axis(2,las=2)
mtext("I",side=2, padj=padj,las=2)
# plot(x, as.vector(y)-x, pch=16, xlab='', ylab='')
# par(new=TRUE)
# plot(x2, as.vector(y2)-x2, pch=16, axes=F, col='red', xlab='',ylab='')
# axis(1, col='red',line=2); 
# axis(4, col='red')
par(mar=c(4,2,2,1), mgp=c(1,0.3,0))
dat = matrix(readBin(dataFiles[32], what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)
fp = min(which(dat[,1] > 100))
ex4 = dat[fp + (-1e2:2e2),]*vpau
matplot(1:nrow(ex4)/fs*1e6-50,ex4, main = 'Example short pulse', type='l', lty=1, xlab='Time (us)', ylab='Amplitude (V)', yaxt='n')
axis(2,las=2)
mtext("J",side=2, padj=padj,las=2, line=1)

@@

#y = ex4[163:220,2]
#x=1:length(y)/2
#plot(x,y)
#mod = nls(y~a*(1-exp(-x/t))+b, start=list(a=-10,t=10,b=5))
#lines(x,predict(mod))
@

@@

\bibliographystyle{nat}
\bibliography{stimjimBibliography}

\begin{table}
\small{
\begin{tabular}{|l|c|c|}
\hline
                                            & Value     & Units     \\
\hline
 Slew rate (voltage mode)                    & 4.1         & V/us     \\
 Slew rate (current mode)                    & 0.8         & V/us     \\
 Output voltage noise (voltage mode)$^a$     & 1           & mV rms    \\
                                             & 5           & mVpp      \\
 Output voltage noise (current mode)$^a$     & 0.6         & mV rms     \\
                                             & 3           & mVpp      \\
 Voltage output step size (voltage mode)$^b$ & 0.45        & V         \\
 Current output step size (current mode)$^b$ & 0.1         & uA        \\ 
 Trigger latency                             & 23          & us        \\
 \hline
\end{tabular}
}
\caption{\label{tab:stimjimSpecs} \textbf{Stimjim electrical characteristics.} $^a$Measured on Tektronix TDS210 digital oscilloscope, full bandwidth (60MHz), 25ms segments.. Note that this is the noise only when the pulse is delivered; at all other times the output is directly connected to ground. Voltage noise in current mode was measured with a 9.75 k$\Omega$ resistive load. $^b$Calculated as output range divided by resolution of output DAC (16 bit).} 
\end{table}

>>>>

\documentclass[twocolumn]{article}
%\documentclass{article}
\usepackage[margin=2cm]{geometry}
%\usepackage[superscript,biblabel]{cite}
\usepackage[super]{natbib}
\usepackage[font={small}]{caption}
\begin{document}
%\SweaveOpts{concordance=TRUE}

@@

    ishe = which(mmand::erode(mmand::erode(isStageHigh>0, c(rep(1,61),rep(0,60))), rep(1,61)) > 0)
    isle = which(mmand::erode(mmand::erode(isStageLow>0, c(rep(1,61),rep(0,60))), rep(1,61)) > 0)
    sh.median = median(x[ishe])*vpau
    sh.mean = mean(x[ishe])*vpau
    sh.sd = sd(x[ishe])*vpau
    sl.median = median(x[isle])*vpau
    sl.mean = mean(x[isle])*vpau
    sl.sd = sd(x[isle])*vpau

    pulseMat = sapply(pulseStartIndices, function(j) x[j+(-100:1000)])
    list(IPI.summary=IPI.summary, IPI.sd=IPI.sd, PW.summary=PW.summary, 
             PW.sd=PW.sd, sh.median=sh.median, sh.sd=sh.sd, 
             sl.median=sl.median, sl.sd=sl.sd, sl.mean=sl.mean, sh.mean=sh.mean, pulseMat=pulseMat)
  })
}

@@

\section*{Abstract}
Electrical stimulation is a simple and powerful tool to perturb and understand the function of neurons and neural circuits. Despite this, devices that can provide precise current or voltage stimulation are expensive and closed-source. Here, we introduce Stimjim, a capable and inexpensive (\$200) open-source instrument for electrical stimulation that combines both function generation and electrical isolation. Stimjim provides microsecond temporal resolution with microampere or millivolt level precision on two electrically isolated output channels. To show the utility of Stimjim, we use it to train mice to perform intracranial self-stimulation (ICSS) for brain stimulation reward (BSR) in a novel head-fixed paradigm. By varying either the output frequency or amplitude, Stimjim enables the experimenter to smoothly tune the strength of reward-seeking behavior.

@@

In contrast, there has been a recent push within the scientific community to produce open labware - open source hardware and software replacements for a variety of common laboratory tasks\cite{pearce_open-source_2013,baden_open_2015} Among many others, examples include software and hardware for:
\begin{itemize}
\item recording or stimulating neurons (e.g. Open Ephys\cite{siegle_open_2017,buccino_open_2018}, Cyclops\cite{newman_optogenetic_2015}, Neurorighter\cite{newman_closed-loop_2013}, Bonsai\cite{lopes_bonsai:_2015})
\item amplifying DNA (e.g., OpenPCR\cite{noauthor_openpcr_nodate}) or preparing single cells for DNA sequencing \cite{booeshaghi_design_2019}
\item imaging microscopic samples (e.g., FlyPi\cite{chagas_100_2017})
\item quantifying chemical reactions or microbial growth in 96-well plates (e.g., \cite{szymula_open-source_2019})
\item growing and evolving bacteria (e.g. \cite{takahashi_low_2015}) 
\item ecological monitoring (e.g., Audiomoth\cite{hill_audiomoth:_2018})
\end{itemize}
We now add Stimjim to this growing body of open hardware. Stimjim replaces \textit{in vivo} neural stimulators at a fraction of the cost, without compromising on quality. Furthermore, due to its entirely open design and software, Stimjim can be modified by end users to fit their specific needs.

@@

<<benchmarkFigure, echo=FALSE,fig.height=8,fig.width=5,fig.cap="\\label{fig:benchmarking} \\textbf{Electrical benchmarks show Stimjim provides microsecond temporal precision and millivolt- and microampere-scale amplitude precision.} \\\\ (A) Example pulse trains with varying inter-pulse interval (IPI). \\\\ (B) Median IPI in a 1-second pulse train measured on a high-speed data acquisition device, plotted against target IPI. Stimjim provides precise frequency control.  \\\\ (C) Standard deviation of inter-pulse intervals measured in a 1-second pulse train. \\\\ (D) Example pulse trains with varying pulse width (PW). \\\\ (E) Median measured PW vs target PW, showing Stimjim has precise pulse width control. \\\\ (F) Standard deviation of pulse widths over 100 pulses. \\\\ (G) Example pulses with varying amplitude.\\\\ (H) Error in amplitude of voltage pulses (100 us, 1 kHz) vs target amplitude.\\\\ (I) same as (H) but for current pulses. \\\\ (J) Example of a fast biphasic (30 us/phase) pulse. Red line is a 500uA current pulse (with a 9.75k resistive load), black line is a 5V pulse. \\\\ (K) Example of a small-amplitude pulse. Red line is a 20uA current pulse (with a 9.75k resistive load), black line is a 0.2V pulse.", fig.align='center', out.width='1\\linewidth', fig.env="figure">>=

@@

plot(paramMat[1:30,3]/1e3, sapply(res[seq(1,60,by=2)], function(x) x$IPI.sd)*1e6, yaxt='n',
  log='x', main='', cex=0.75, xlab='Target IPI (ms)', ylab='IPI standard deviation (us)')
points(paramMat[1:30,3]/1e3, sapply(res[seq(2,60,by=2)], function(x) x$IPI.sd)*1e6,
  col='red', cex=0.75)
axis(2,las=2)
mtext("C",side=2, padj=padj,las=2)

@@

plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) x$PW.sd)*1e6/2, yaxt='n',
  log='x', main='', xlab='Target PW (us)', ylab='PW standard deviation (us)', cex=0.75)
points(paramMat[31:59,7], sapply(res[seq(62,118,by=2)], function(x) x$PW.sd)*1e6/2,
  col='red', cex=0.75)
axis(2,las=2)
mtext("F",side=2, padj=padj,las=2)

ex3 = sapply(dataFiles[seq(81,61,length.out=6)], function(f){
  dat = matrix(readBin(f, what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)
  fp = min(which(dat[,1] > 10))
  dat[fp + (-200:700),1]*vpau
})
par(mar=c(4,1,2,1))
matplot(1e3/fs*1:nrow(ex3), dendRites::offset(ex3,15), col=1, type='l', lty=1, bty='n', yaxt='n', ylab='', xlab='time (ms)',
                main = "Varying amplitude");
mtext("G",side=2, padj=padj,las=2)

par(mar=c(4,3,2,1))
x = c(paramMat[61:89,5], -paramMat[61:89,5])
y = t(sapply(res[seq(121,178,by=2)], function(x) c(x$sh.mean,x$sl.mean))*1e3)
plot(x/1000,y-x, main='Voltage accuracy', xlab='Target amplitude (V)', ylab='Measured error (mV)', cex=0.75, yaxt='n', ylim=c(-30,30))
axis(2,las=2)
mtext("H",side=2, padj=padj,las=2)

x2 = c(paramMat[61:90,6], -paramMat[61:90,6])
y2 = t(sapply(res[seq(122,180,by=2)], function(x) c(x$sh.mean,x$sl.mean))/9750*1e6)
plot(x2/1000,y2-x2, main='Current accuracy', xlab='Target amplitude (mA)', ylab='Measured error (uA)', cex=0.75, col='red', yaxt='n',ylim=c(-5,5))
axis(2,las=2)
mtext("I",side=2, padj=padj,las=2)

par(mar=c(4,2,2,1), mgp=c(1,0.3,0))
dat = matrix(readBin(dataFiles[32], what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)
fp = min(which(dat[,1] > 100))
ex4 = dat[fp + (-1e2:2e2),]*vpau
matplot(1:nrow(ex4)/fs*1e6-50,ex4, main = 'Example short pulse', type='l', lty=1, xlab='Time (us)', ylab='Amplitude (V)', yaxt='n')
axis(2,las=2)
mtext("J",side=2, padj=padj,las=2, line=1)

@@

@@

\bibliographystyle{unsrt}
\bibliography{stimjimBibliography}

\begin{table}
\small{
\begin{tabular}{|l|c|c|}
\hline
                                            & Value     & Units     \\
\hline
 Slew rate (voltage mode)                    & 4.1         & V/us     \\
 Slew rate (current mode)                    & 0.8         & V/us     \\
 Output voltage noise (voltage mode)$^a$     & 1           & mV rms    \\
                                             & 5           & mVpp      \\
 Output voltage noise (current mode)$^a$     & 0.6         & mV rms     \\
                                             & 3           & mVpp      \\
 Voltage output step size (voltage mode)$^b$ & 0.45        & V         \\
 Current output step size (current mode)$^b$ & 0.1         & uA        \\ 
 Output impedance (voltage mode)$^c$         & 180         & $\Omega$ \\ 
 Trigger latency                             & 23          & us        \\
 \hline
\end{tabular}
}
\caption{\label{tab:stimjimSpecs} \textbf{Stimjim electrical characteristics.} $^a$Measured on Tektronix TDS210 digital oscilloscope, full bandwidth (60MHz), 25ms segments.. Note that this is the noise only when the pulse is delivered; at all other times the output is directly connected to ground. Voltage noise in current mode was measured with a 9.75 k$\Omega$ resistive load. $^b$Calculated as output range divided by resolution of output DAC (16 bit). $^c$ value from from datasheet for Vishay DG509B (output switch). } 
\end{table}
